Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-30 ThesisJP/Background/background.tex
Revised GPU text.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufactures of central processing units (CPUs) hit a frequency-wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.  From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end, and where shaded a rasterized geometry was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Secondly, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is shaded, projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU and serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array (processor)}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics, hence they can be classified as APUs. An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer, it is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and a GPU.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}), which has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32, 32 bit, floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floating point values per clock cycle at 745 MHz\footnote{Nvidia typically refers to a core as a unit capable of performing a FMA instruction on one single precision floating point value. Hence, in Nvidia's world, Kepler has $15*192=2880$ cores.}. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of five TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent. Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five in raw computationally power. A claimed speed-up of several order of magnitude is typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}. In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications and differences in theoretical peak performance.

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the graphics pipeline. When GPGPU gained interest and the GPU was used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which are maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. Where CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. An example of a CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the size of the buffer \texttt{a}.

When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block. Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not depend on the output from other blocks, hence, it must be possible for the GPU to schedule blocks in any order.

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as unit capable of processing one data element in the SIMD instruction. This is why GPU threads are said to be light weighted. This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions when programing for the GPU. This often leads to a highly optimized parallel implementation on the GPU and not on the CPU. Threads, or SIMD instructions, is just how one program for the GPU, there is no other way of doing it.

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase the only algorithms that will benefit is those who can be divided into enough concurrent fine granular parts. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by issuing instructions. All of this is something which is impossible to achieve in real life applications. One will always end up with a throughput which is a fraction of the peak throughput.

The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing. A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in near future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating complexity of BS-Capon beamformer \textbf{Paper II} we find that is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.

@@

For the simulator presented in \textbf{Paper V} the required number of FLOPS in order to achieve a certain frame rate $f_{\text{FPS}}$ is:
\begin{align}
\text{FLOPS}_{\text{HOS}} = N_{\text{source points}}N_{\text{field points}}f_{\text{FPS}}.
\end{align}
Therefore, to simulate a field of one million points with several thousand source points at 50 FPS, more than 100 GFLOPS is required. Note that this simulator evaluates a cosine and sine function for each calculation, something which makes it even harder to achieve the required throughput, since especially GPUs have less silicon devoted to such tasks.

>>>>

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate until the required cooling finally became to much in 2004.  From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and colors where fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.

The graphics pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advanced feature found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the problem of rendering graphics. When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem. How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to unify the computational power of both processors. Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array (processor)}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processor is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics, and all of them are therefore APUs. An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers. An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a high level of CPU-GPU data synchronization.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU. This is a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is designed to be the main control unit of a personal computer, and is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these features are and why there is still a big difference in peak performance between a CPU and a GPU.

@@

In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions. These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS}.
\end{equation}
Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture by Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15*192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that the actual throughput is highly algorithm dependent. Nevertheless, in theory, the computational power provided by modern GPUs is five times higher than that provided by CPUs. A claimed speed-up of several orders of magnitude is therefore typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}. However, in the next section we will see why it is often easier to obtain an optimized implementation for the GPU than for the CPU.

In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications and differences in theoretical peak performance.

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the graphics pipeline. When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which is maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. While CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

Both CUDA and OpenCL are centered around a kernel function which is executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. An example of a CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA this kernel function is launched by using a CUDA-specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} are 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the size of the buffer \texttt{a}.

When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which values to gather from the global memory, in this case an element of \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel are finished for all threads in the block. Thus,  in order to formulate a given problem as a kernel function, the global memory input to each thread must not depend on the output from other blocks. Hence, it must be possible for the GPU to schedule blocks in any order.

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction. The GPU is because of this sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor. This also explain why GPU threads are said to be light-weighted. Such a conceptual difference means that there is an inherent focus on breaking algorithms down into SIMD instructions when programing for the GPU. This often leads to a highly optimized parallel implementation on the GPU and not on the CPU. Manually typed SIMT, or SIMD, instructions is the only way of programming the GPU, while a single thread with automatic vectorization of for-loops is the standard way of programming the CPU.

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into enough independent calculations. Also, to achieve the peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something which is impossible to achieve in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.

The code in Fig.\,\ref{fig:gpu_grid} is an incomplete and minimal example of a GPU program. A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is part of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Therefore, we will probably see a lot of work aimed at reducing this overhead in the future.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in near future, programming the GPU will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction. The relevance of auto-applied GPU instructions will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, GPU programming will probably be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.

@@

For the simulator presented in \textbf{Paper V} the required number of FLOPS in order to achieve a certain frame rate $f_{\text{FPS}}$ is:
\begin{align}
\text{FLOPS}_{\text{HOS}} = N_{\text{source points}}N_{\text{field points}}f_{\text{FPS}}.
\end{align}
Therefore, to simulate a field of one million points with several thousand source points at 50 FPS, more than 100 GFLOPS is required. Note that this simulator evaluates a cosine and sine function for each calculation, something which makes it even harder to achieve the required throughput, since the GPU has less silicon devoted to such tasks (SFUs).
